Title of the report

### 2011-12-05 doc/report/report.tex
dependent types, comments on report

\section*{Acknowledgements}
% Ramona, Aarne, Elisabet, Markus, Peter?, Lars, Dan, Krasimir etc

\newpage
\chapter{Introduction}
\section{Introduction} 
Ultimate goal: want computers to 'understand' language. Not good with flat strings.
We need to translate strings of natural language to a richer sturture,
a parse tree. So far there is no
freely availbe ruled-based parsing giving a deep analyse for Swedish.
%There are parsers which put only put part-of-speach tags to
%the text, and others that gives a deeper analyse. 
There are statistical parser, and parser which only are deal with controlled language.
The GF parser are one of those. 
%What is parsing, flat strings to trees, example picture, what is deep parsing.
So far stable and good when working with controlled languges, but recent experiments
have showed that it is also possible to parse free language.
To do this for Swedish, we need to  .. the resource grammar needed to covering
the language specific constructions. Further a lagre lexicon is also needed, as well
as a method for evaluation.
During this project, we have worked on and explored those things and some problems identified.
We have extended the Swedish resource grammar in GF, to give a better coverage and imported
a large lexicon. Work to compare the trees from Talbanken to GF and to translate them.

@@

\section{Background}  
To build a parser or a grammar from scratch is a heavy work. The idea for this
project is to combine already existing parts into a bigger and more expressive grammar
as well as a large scale GF lexicon.
As all GF grammars, this one can be used for parsing, and we develop it by getting
examples, ideas and test material from a treebank.
The project is hence heavily dependening on three resources, which will be described
in the following sections.

\subsection{Swedish}
Swedish\cite[Inl. \textsection 3]{SAG}is a North-Germanic language,
closely related to Norweigan and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also 
an official language in Finland and spoken by approximately 10 million people.
Swedish syntax are often similair to English, but have some characteristics:
\subsubsection*{Verb second}
Swedish is a verb-second language\cite[p. 116]{gunlog}: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be fronted\cite[\textsection 1027]{H&H}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being emphasized.
\enumsentence{
\shortex{4}
{Du & ser & mig & inte.}
{You & see & me & not.}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{Me & see & you & not.}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{Not&  see& you& me.}} \label{ex:swedish-avso}

@@

\subsection{Saldo}
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The words are equipped with both sematic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of sematically connected words.\\
\newpage % only until saldo can be on same page as the pic
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

If one word
have different meanings, there will be separate entries for each of them.
The word ``växa" (``grow") for example, has two meanings: \emph{getting bigger}
\enumsentence{Gräset växte varje dag\\
The grass grew each day}
or \emph{being a plant}. 
\enumsentence{Det växte en gammal ek där.\\
There stood an old oak.}
The word ``växa" is however declined the same way for both interpretations of the word.
Hence Saldo's morpological lexicon, provided under LPGL in XML format,
only keeps one entry of ``växa".
For the homonyms that does not share the whole inflection tables, different entries are 
kept in the morphological lexicon.

For our purpose, only the morphological lexcion was needed.
The data can be processed and translated to GF format as described in 
section \ref{sec:prog.saldo}.

\subsection{Grammatical Framework}
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a functional grammar formalism based on Martin-Löf type theory.
It has been developed for about 20? years and are inspired by programming langauges
like Haskell\cite{haskell},
ML\cite{ml}, Prolog\cite{bratko} and
Agda\cite{agda}.
Grammars written in GF can be used together with languages 
like Java, Java script and Haskell and works for both parsing and generation.

An important feature of GF is
%Being developed with multilinguality in mind, GF makes an important 
the distiction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
shared by many languages, it models the semantics of the domain without concern for
language specific features such as agreement or word order.

@@

Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|,
and \verb|Pred| simply glues the two strings ``katten" and ``sover" together:
\verb|Pred cat sleep = "katten sover"|.\\
We get a more complicated example if we allow the nouns to be used in both
plural and singular. To do this, we add a category \verb|NP| to the
abstract. \verb|N| now means a noun
in any of the number, and we introduce two funciton, \verb|NSg : N -> NP| and
\verb|NPl : N -> Pl| which sets the number.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; NP ; V ; S ;               lincat V, S, NP = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : NP -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> NP ;                    NPl n = n.s ! Pl ;
    NPl : N -> NP ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records,tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s| shows that it is a table, which given a parameter of type \verb|Num| returns
a string. \verb|Num| is defined as a parameter, which can either have value \verb|Sg|
or \verb|Pl|. In \verb|NPl| and \verb|NSg|, \verb|n.s| means that we use the
field \verb|s| of \verb|n|, and then the selection operator \verb|!| is used to
select a branch in the table.

@@

A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are WebAlt\cite{webalt} and something else.
%, the Online Syllogism Solver\cite{malin!}. 
The framework is also used in the
european project MOLTO\cite{molto} for online translation between upto 15 language.\\

\subsubsection{The resource library}
The resource library\cite{gf-resource} is a very useful part of the framework. It is provided in the 
GF package. %, they give a good start for all who want to write an application grammar.
Instead of letting the users of the framework reimplement the morphological and
the fundamental syntactic structures of a language for each application, all
this can be found as libraries. So far there are 20 languages in the library, and
they all share the same abstract syntax.
There are also a small test lexicon included, containing some hundered of common words.

The resource grammars describe how to construct phrases and sentences and how to
decline words. The latter is done by smart paradigms: functions that analyses
some given forms of a word to find out how the inflection table should look.
For example, the declination of many Swedish nouns can be determined by looking
at the singular form only. 
\begin{wrapfigure}{l}{0.4\textwidth}
\begin{tabular}{| l | l |}
\hline
1st declination & 5th declination \\
\hline
flick\textbf{a}    &     hjärt\textbf{a}   \\
flick\textbf{an}    &    hjärt\textbf{at}  \\
flic\textbf{kor}    &    hjärt\textbf{an}  \\
flic\textbf{korna}  &    hjärt\textbf{ana} \\
\hline
\end{tabular}
\end{wrapfigure}
This is the case for ``flicka", the correct declination
can be guessed by using the first declination. % and noting that the word ends with an ``a". 
%Like most Swedish nouns ending with an ``a", it follows the first declination.
For others, like "hjärta", also the plural form "hjärtan" is needed.
%, even though it ends with an ``a". 
The worst case is nouns that need four forms, both singular and
plural in definite and indefinite form.

\vspace{5mm}
As an open source-project, GF is constantely being developed and improved. New
languges are added, the compiler is being improved, ways of använad det på ett
lätt sätt are added and the possiblities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for solving syllogism or generating natural language via Montague
semantics.
Further, experiments on using GF for free parsing has been conducted \cite[\textsection 4.1]{gfMech} 
This is also a future goal for our project and one of the delmål has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

@@

\subsection{Talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being an freely avaible, manually annotade, large scale treebank,
Talbanken\cite{talbanken} was perfect for our purpose.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and the last one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{malt}
and was thus redistributed in a more modern version\cite{talbanken05} This work
was done by Nivre, Hall and Nilsson in 2005. They released it in Malt and Tiger XML-formats,
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Related work}
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt\cite{malt} ... .
However has been two, about CLE\cite{cle} and what happend to it. Logic and translation
to Swedish, The spoken lang. translator\cite{cle2}. Wiren \cite{wiren}, cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

@@

\section{Importing Saldo}
\label{sec:prog.saldo}
Saldo is compatible with GF, which makes it easy to transform the morphological
lexicon into our format. The lexicon has been imported earlier (Krasimir,2009?)
but the code needed updates to not miss many important words.
Experiments have been made to use a interactive tool for lexical aquisation,
but the dictionary was still much to small. The aquisation tool may still
be used for complenting the lexicon.

The basic algorithm for porting Saldo is the same as Angelov's, it works by
making use of the smart paradigms in GF.
%Find the paper!
%To understand it, we first need to look at differences between Saldo and a GF lexicon.
The program will try to create a lexicon by finding a number of forms for every
word and feeding them to the smart paradigms. As few forms as possible should be
used, and the resulting table generated by GF should be equivalent to the one found
in Saldo. This process is done iteratively,
starting from the easiest case. For each attempt the results are compared and when
the correct way of using a word has been found, the code is saved.
Figure \ref{pic:TabVax} shows a the information given by Saldo and by GF respectively.\\
\begin{figure}[h]
  \begin{center}
\subfloat[Saldo]{\label{pic:saldoVax}\includegraphics[height=90mm]{saldoVax.png}}
\hspace{5mm}
\subfloat[GF]{\label{pic:gfVax}\includegraphics[height=60mm]{gfVax.png}}
\caption{Inflection table for the verb ``växa" (grow)}
\label{pic:TabVax}
  \end{center}
\end{figure}
The tables does not overlap completely, there are some more forms in Saldo's table (eg. the 
compounding form ``väx-")
while the one generated in GF contains some that are not in Saldo (eg. ``vuxits").
Saldo may also contain variats: ``växt" and ``vuxit" are both supine form.
As far as possible, the program makes up for this by only comparing the overlapping forms
and only requiring that GF generatens one variant whenever alternatives are given. 
The program continues to generate new lexicons as long as there are more
paradigms to choose from and compare the results after each run.

The program has been made more robust than the previous version, it prints
logs and finally lists of the words that were not imported. The logs provides
information about which paradigms that were used for each word and the result
from comparing the inflection tables.
During the porting of Saldo, some renaming of the lemmas was needed.
Special characters like \emph{å,ä,ö} should not occure in GF function names.
Those are translated into \emph{aa,ae,oe} respectively. 
The Saldo identifier \verb|äta..vb| is thus turned into the GF identifier \verb|aeta_V|.
Since there might be two lemmas that share the same
identifier after the translation, a number is added to the name whenever a translation
have been done: 
\begin{tabular}{lll}
kältisk & $\rightarrow$ & kaeltisk\_1 \\
kaeltisk & $\rightarrow$ & kaeltisk \\
entrecôte &$\rightarrow$ & entrecoote\_1 \\ 
\end{tabular}

@@

\subsection{Results}
All words from Saldo were not be imported. There are a number of reasons
for this, the most obvious one is that we do not want all categories from
Saldo in the GF lexicon. Prepositions(\emph{pp}) and numerals(\emph{nl}) etc. are
assumed to be known already,
and are usually not kept in the standard lexicon, but in modules for the basic
structures of the language. Categories involving several words(\emph{vbm,nnm})
need to be handeld differently, usually as idioms. Pronouns(\emph{pn}) are not  
analysed the same way, see section \ref{sec:swegf}. Some experiments on finding
the right GF-category for those have been done, see \ref{sec:gf.quant}.
Six categories were imported: \\
\begin{tabular}{l|lll}
& Saldo tag & GF category & Example \\
\hline
 Adverb & \textbf{ab} &\textbf{Adv} & ofta \\
 Adjective&\textbf{av} &    \textbf{A} & gul\\
 Noun & \textbf{nn} &\textbf{N} & hus\\
 Verb & \textbf{vb} &\textbf{V} & springa\\
 Reflexive verbs  &\textbf{vbm}& \textbf{V2} & raka sig\\
 Verbs with particles &\textbf{vbm}& \textbf{V}  & hitta på\\
\end{tabular}\\

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
One reason for why words of the categories listed above could not be imported
is that Saldo does not contain information about word forms that are not used.
In some cases those forms are needed by the smart paradigm. 
This was a common reason why nouns could not be imported. 
Consider for example
the plural tantum %\cite[44]{SAG}
noun ``glasögon" (``glasses"). Since the smart paradigm
requires a singular form, this lemma could not be added to the lexicon. ``Glasögon" could
be added manually, by using the somewhat incorrect singular form ``glasöga" (``glass-eye").
The same problem occured for verbs that always are used in passive form:
``andas" (``breathe"), ``frodas" (``flourish"), ``lyckas" (``succeeed"). This
was the case for close to 80\% of the failing verbs of type \verb_vb_.

In some cases the smart paradigms could not generate the correct declination,
although this happend seldom. % frysa,fisa passive preterium

@@

\enumsentence{\shortexnt{6}{Han & ser &katten & han  &tycker & om}
              {He & sees & the cat & he & likes &}}
              %{`She sees the sleeping boy'}}
In addition to the basic resource grammar, additional constructions had been added
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
%For example, the \verb|Extra| module included
Among those were functions for topologication, although not all
of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
For example, \verb|FocObj| fronted the object as in sentence \ref{gfSwe:apple}.
\enumsentence{\shortex{6}
{Det & äpplet & vill & jag & inte & ha}
{That & apple & want & I & not &}
{I don't want that apple}}\label{gfSwe:apple}
The module provided functions for preposition stranding:
\enumsentence{\begin{tabular}{ll}
Stranded preposition & cf.\\
Vem måste jag akta mig för? & För vem måste jag akta mig?\\
Who do I need to watch out for? & For whom do I need to wath out?\\
\end{tabular}}

@@

Before describing what has been added, will give a short introductions to the
the cats of resource grammar and the sw ipmlementation in particular.

>>>>

\section*{Acknowledgments}
% Ramona, Aarne, Elisabet, Markus, Peter?, Lars, Dan, Krasimir etc
% Funding!

% check references! lncs is good maybe. lnai
\newpage
\chapter{Introduction}
\section{Introduction} 
>>This work is part of bigger project, spring and next spring.
>>Connection between grammar & parser: GF, every grammar defines a parser.
Ultimate goal: want computers to 'understand' language. Not good with flat strings.
We need to translate strings of natural language to a richer structure,
a parse tree. So far there is no
freely available ruled-based parsing giving a deep analyse for Swedish.
%There are parsers which put only put part-of-speach tags to
%the text, and others that gives a deeper analyse. 
There are statistical parser, and parser which only are deal with controlled language.
The GF parser are one of those. 
%What is parsing, flat strings to trees, example picture, what is deep parsing.
So far stable and good when working with controlled languages, but recent experiments
have showed that it is also possible to parse free language.
To do this for Swedish, we need to  .. the resource grammar needed to covering
the language specific constructions. Further a large lexicon is also needed, as well
as a method for evaluation.
During this project, we have worked on and explored those things and some problems identified.
We have extended the Swedish resource grammar in GF, to give a better coverage and imported
a large lexicon. Work to compare the trees from Talbanken to GF and to translate them.

@@

\section{Background}  
>>This work is part of bigger project, spring and next spring.
>>Connection between grammar & parser: GF, every grammar defines a parser.
To build a parser or a grammar from scratch is a heavy work. The idea for this
project is to combine already existing parts into a bigger and more expressive grammar
as well as a large scale GF lexicon.
As all GF grammars, this one can be used for parsing, and we develop it by getting
examples, ideas and test material from a treebank.
The project is hence heavily depending on three resources, which will be described
in the following sections.

\subsection{Swedish}
Swedish\cite[Inl. \textsection 3]{SAG}is a North-Germanic language,
closely related to Norwegian and Danish. The languages share most of their
grammatical structures and are mutual intelligibility. Swedish is also 
an official language in Finland and spoken by approximately 10 million people.
Swedish syntax are often similar to English, but have some characteristics:
\subsubsection*{Verb second}
Swedish is a verb-second language\cite[p. 116]{gunlog}: the
second constituent of a declarative main clause must consist of a verb.
The normal word order is subject-verb-object, but any syntactic category can be fronted\cite[\textsection 1027]{H&H}.
Topicalisation is very common, especially for temporal and spatial adverbial phrases.
The examples \ref{ex:swedish-svo} - \ref{ex:swedish-avso} all have the same propositional
meaning, although the fronted part of sentence \ref{ex:swedish-ovs} and \ref{ex:swedish-avso} are being emphasized.
\enumsentence{
\shortex{4}
{Du & ser & mig & inte.}
{You & see & me & not.}
{`You  don't see me'.}} \label{ex:swedish-svo}
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Mig  & ser &  du & inte.}
{Me & see & you & not.}}\label{ex:swedish-ovs} 
\vspace{-3mm}
\enumsentence{
\shortexnt{4}
{Inte & ser& du &mig.}
{Not&  see& you& me.}} \label{ex:swedish-avso}

@@

\subsection{Saldo}
>> syntactic information too. Mention that GF does not have this, only lexical or
>> semantic. Not 1-1.
A good parser needs a good lexicon. We have used of Saldo\cite{saldo}, a
large electronic lexicon developed and maintained at Gothenburg University. It is
built upon Svenskt Associations Lexicon and contains information about more than 
120 000 modern Swedish words.
The words are equipped with both semantic and a morphological
information. The user can find compounding analyses, examples of usage in corpora,
and graphs of semantically connected words.\\
\newpage % only until saldo can be on same page as the pic
\begin{figure}[h!]
\centering
\subfloat[Morphological info for ``katt"]{\label{pic:saldoTab}\includegraphics[width=40mm]{saldotab.png}}
\hspace{20mm}
\subfloat[Graph showing the hyponyms of ``katt"]{\label{pic:saldoWheel}\includegraphics[width=70mm]{saldograph.png}}
\caption{Information in Saldo}
\label{fig:saldo}
\end{figure}

If one word
have different meanings, there will be separate entries for each of them.
The word ``växa" (``grow") for example, has two meanings: \emph{getting bigger}
\enumsentence{Gräset växte varje dag\\
The grass grew each day}
or \emph{being a plant}. 
\enumsentence{Det växte en gammal ek där.\\
There stood an old oak.}
The word ``växa" is however declined the same way for both interpretations of the word.
Hence Saldo's morphological lexicon, provided under LPGL in XML format,
only keeps one entry of ``växa".
For the homonyms that does not share the whole inflection tables, different entries are 
kept in the morphological lexicon.

For our purpose, only the morphological lexicon was needed.
The data can be processed and translated to GF format as described in 
section \ref{sec:prog.saldo}.

\subsection{Grammatical Framework}
>> more here. First: grammar formalism with Martin-Löf.
>> snd: programming language, type checking and logic: prolog (check this)
>> dependent and implicit types : agda. (read gf book and krasses book)
>> pgf\ref{}, runtime system with parsing, lin, random generation osv.
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a functional grammar formalism based on Martin-Löf type theory.
It has been developed for about 20? years and are inspired by programming languages
like Haskell\cite{haskell},
ML\cite{ml}, Prolog\cite{bratko} and
Agda\cite{agda}.
Grammars written in GF can be used together with languages 
like Java, Java script and Haskell and works for both parsing and generation.

>> The most important!!
An important feature of GF is
%Being developed with multilinguality in mind, GF makes an important 
the distinction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
shared by many languages, it models the semantics of the domain without concern for
language specific features such as agreement or word order.

@@

Figure \ref{fig:gfSweCnc1} shows how the abstract grammar can be implemented
for Swedish. Nouns, verbs and sentences are all defined as strings, \verb|Str|,
and \verb|Pred| simply glues the two strings ``katten" and ``sover" together:
\verb|Pred cat sleep = "katten sover"|.\\
We get a more complicated example if we allow the nouns to be used in both
plural and singular. To do this, we add a category \verb|NP| to the
abstract. \verb|N| now means a noun
in any of the number, and we introduce two function, \verb|NSg : N -> NP| and
\verb|NPl : N -> NP| which sets the number.
\begin{figure}[h!]
\begin{verbatim}              
 abstract TestGrammar = {          concrete TestGrammarSwe of TestGrammar = {
  cat N ; NP ; V ; S ;               lincat V, S, NP = Str ;
                                            N = {s : Num => Str} ;
  fun                                lin   
    Pred : NP -> V -> S ;              Pred n v = n ++ v ;
    NSg : N -> NP ;                    NPl n = n.s ! Pl ;
    NPl : N -> NP ;                    NSg n = n.s ! Sg ;
    cat_N : N ;                        cat_N = {s = table {Sg => "katten" ;
    sleep_V : V ;                                          Pl => "katterna"}};
 }                                     sleep_V = "sover"  ;
                                     param Num = Sg | Pl ;
                                     }
\end{verbatim}           
\caption{A modified grammar}
\label{fig:gfTest2}
\end{figure}
Figure \ref{fig:gfTest2} introduces some new concepts: records,tables and parameters.
In the concrete syntax, \verb|N| is defined to be a record consisting of the field
\verb|s|. The type of \verb|s| shows that it is a table, which given a parameter of type \verb|Num| returns
a string. \verb|Num| is defined as a parameter, which can either have value \verb|Sg|
or \verb|Pl|. In \verb|NPl| and \verb|NSg|, \verb|n.s| means that we use the
field \verb|s| of \verb|n|, and then the selection operator \verb|!| is used to
select a branch in the table.

@@

>>not very restricted, not few ambiguities.
>>the scope is controlled language, but possible ambiguities (discuss this in future
>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
>>syntactical (man with the telescope), 
>>lexical (youSg/Pl, växa).
A GF grammar usually covers a very restricted set of language where few ambiguities
occur and much care can be taken to preserving the semantics of the
different languages.
By using a controlled language like this,  good and reliable translation can be
ensured. Some examples of projects using GF are WebAlt\cite{webalt} and something else.
>> key, talk, hats, bom (see references in Ramonas thesis).
%, the Online Syllogism Solver\cite{malin!}. 
The framework is also used in the
European project MOLTO\cite{molto} for online translation between up to 15 language.\\

\subsubsection{The resource library}
>>Best part? Biggest library?
>>Most general resource, first large grammar, covers fundamental syntactics.
The resource library\cite{gf-resource} is a very useful part of the framework. It is provided in the 
GF package. %, they give a good start for all who want to write an application grammar.
Instead of letting the users of the framework reimplement the morphological and
the fundamental syntactic structures of a language for each application, all
this can be found as libraries. So far there are 20 languages in the library, and
they all share the same abstract syntax.
There are also a small test lexicon included, containing some hundred of common words.

The resource grammars describe how to construct phrases and sentences and how to
decline words. The latter is done by smart paradigms: functions that analyses
some given forms of a word to find out how the inflection table should look.
For example, the declination of many Swedish nouns can be determined by looking
at the singular form only. 
\begin{wrapfigure}{l}{0.4\textwidth}
\begin{tabular}{| l | l |}
\hline
1st declination & 5th declination \\
\hline
flick\textbf{a}    &     hjärt\textbf{a}   \\
flick\textbf{an}    &    hjärt\textbf{at}  \\
flick\textbf{or}    &    hjärt\textbf{an}  \\
flick\textbf{orna}  &    hjärt\textbf{ana} \\
\hline
\end{tabular}
\end{wrapfigure}
This is the case for ``flicka", the correct declination
can be guessed by using the first declination. % and noting that the word ends with an ``a". 
%Like most Swedish nouns ending with an ``a", it follows the first declination.
For others, like "hjärta", also the plural form "hjärtan" is needed.
%, even though it ends with an ``a". 
The worst case is nouns that need four forms, both singular and
plural in definite and indefinite form.

\vspace{5mm}
>> references here! to everything. Remove uncertain, add ontology, aarne paper about
>> donkey, find aarnes paper in ramona's thesis (french), hats formal methods,
>> bom.
As an open source-project, GF is constantly being developed and improved. New
languages are added, the compiler is being improved, ways of använad det på ett
lätt sätt are added and the possibilities to use GF in different environments
increased. There are  research about how to make better use of the dependent
types, for solving syllogism or generating natural language via Montague
semantics.
Further, experiments on using GF for free parsing has been conducted \cite[\textsection 4.1]{gfMech} 
This is also a future goal for our project and one of the delmål has been to
model an important part of Swedish in GF. The starting point was the resource
grammar and the result should still have a strong connection to this.

@@

\subsection{Talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being an freely available, manually annotated, large scale treebank,
Talbanken\cite{talbanken} was perfect for our purpose.
It was created in the 1970s at Lund University. In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and the last one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser\cite{malt}
and was thus redistributed in a more modern version\cite{talbanken05} This work
was done by Nivre, Hall and Nilsson in 2005. They released it in Malt and Tiger XML-formats,
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

\section{Related work}
>> Much more here, important!! comparisons.
Computational ling is active area for Swedish, there are many other Swedish parser,
but grammars?
Many of the parsers not ruled based, the most known one Malt\cite{malt} ... .
However has been two, about CLE\cite{cle} and what happened to it. Logic and translation
to Swedish, The spoken lang. translator\cite{cle2}. Wiren \cite{wiren}, cassSwe \cite{casswe}
Other Sw. parsers and grammars, evaluation of them? 
Other grammar frameworks.
% related and references work: CLE, Krasimirs book, CassSwe, GF book,SAG, gunlög,
% the multilingual treebank, Homse&Hinchcliff, Aarne i LILT

@@

\section{Importing Saldo}
\label{sec:prog.saldo}
Saldo is compatible with GF, which makes it easy to transform the morphological
lexicon into our format. The lexicon has been imported earlier (Krasimir,2009?)
but the code needed updates to not miss many important words.
Experiments have been made to use a interactive tool for lexical acquisition,
but the dictionary was still much to small. The acquisition tool may still
be used for complementing the lexicon.

The basic algorithm for porting Saldo is the same as Angelov's, it works by
making use of the smart paradigms in GF.
%Find the paper!
%To understand it, we first need to look at differences between Saldo and a GF lexicon.
The program will try to create a lexicon by finding a number of forms for every
word and feeding them to the smart paradigms. As few forms as possible should be
used, and the resulting table generated by GF should be equivalent to the one found
in Saldo. This process is done iteratively,
starting from the easiest case. For each attempt the results are compared and when
the correct way of using a word has been found, the code is saved.
Figure \ref{pic:TabVax} shows a the information given by Saldo and by GF respectively.\\
\begin{figure}[h]
  \begin{center}
\subfloat[Saldo]{\label{pic:saldoVax}\includegraphics[height=90mm]{saldoVax.png}}
\hspace{5mm}
\subfloat[GF]{\label{pic:gfVax}\includegraphics[height=60mm]{gfVax.png}}
\caption{Inflection table for the verb ``växa" (grow)}
\label{pic:TabVax}
  \end{center}
\end{figure}
The tables does not overlap completely, there are some more forms in Saldo's table (eg. the 
compounding form ``väx-")
while the one generated in GF contains some that are not in Saldo (eg. ``vuxits").
Saldo may also contain variats: ``växt" and ``vuxit" are both supine form.
As far as possible, the program makes up for this by only comparing the overlapping forms
and only requiring that GF generates one variant whenever alternatives are given. 
The program continues to generate new lexicons as long as there are more
paradigms to choose from and compare the results after each run.

The program has been made more robust than the previous version, it prints
logs and finally lists of the words that were not imported. The logs provides
information about which paradigms that were used for each word and the result
from comparing the inflection tables.
During the porting of Saldo, some renaming of the lemmas was needed.
Special characters like \emph{å,ä,ö} should not occur in GF function names.
Those are translated into \emph{aa,ae,oe} respectively. 
The Saldo identifier \verb|äta..vb| is thus turned into the GF identifier \verb|aeta_V|.
Since there might be two lemmas that share the same
identifier after the translation, a number is added to the name whenever a translation
have been done: 
\begin{tabular}{lll}
kältisk & $\rightarrow$ & kaeltisk\_1 \\
kaeltisk & $\rightarrow$ & kaeltisk \\
entrecôte &$\rightarrow$ & entrecoote\_1 \\ 
\end{tabular}

@@

\subsection{Results}
All words from Saldo were not be imported. There are a number of reasons
for this, the most obvious one is that we do not want all categories from
Saldo in the GF lexicon. Prepositions(\emph{pp}) and numerals(\emph{nl}) etc. are
assumed to be known already,
and are usually not kept in the standard lexicon, but in modules for the basic
structures of the language. Categories involving several words(\emph{vbm,nnm})
need to be handled differently, usually as idioms. Pronouns(\emph{pn}) are not  
analysed the same way, see section \ref{sec:swegf}. Some experiments on finding
the right GF-category for those have been done, see \ref{sec:gf.quant}.
Six categories were imported: \\
\begin{tabular}{l|lll}
& Saldo tag & GF category & Example \\
\hline
 Adverb & \textbf{ab} &\textbf{Adv} & ofta \\
 Adjective&\textbf{av} &    \textbf{A} & gul\\
 Noun & \textbf{nn} &\textbf{N} & hus\\
 Verb & \textbf{vb} &\textbf{V} & springa\\
 Reflexive verbs  &\textbf{vbm}& \textbf{V} & raka sig\\
 Verbs with particles &\textbf{vbm}& \textbf{V}  &  piggna till\\
\end{tabular}\\

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
One reason for why words of the categories listed above could not be imported
is that Saldo does not contain information about word forms that are not used.
In some cases those forms are needed by the smart paradigm. 
This was a common reason why nouns could not be imported. 
Consider for example
the plural tantum %\cite[44]{SAG}
noun ``glasögon" (``glasses"). Since the smart paradigm
requires a singular form, this lemma could not be added to the lexicon. ``Glasögon" could
be added manually, by using the somewhat incorrect singular form ``glasöga" (``glass-eye").
The same problem occurred for verbs that always are used in passive form:
``andas" (``breathe"), ``frodas" (``flourish"), ``lyckas" (``succeeed"). This
was the case for close to 80\% of the failing verbs of type \verb_vb_.

In some cases the smart paradigms could not generate the correct declination,
although this happened seldom. % frysa,fisa passive preterium

@@

\enumsentence{\shortexnt{6}{Han & ser &katten & han  &tycker & om}
              {He & sees & the cat & he & likes &}}
              %{`She sees the sleeping boy'}}
In addition to the basic resource grammar, additional constructions had been added
to the module \verb|Extra|. The functions given here do not have to be translatable
to all other language, but are meant to cover language specific constructions.
%For example, the \verb|Extra| module included
Among those were functions for topologication, although not all
of them were implemented for Swedish.
%word order, agreement, tempus, adverbs, adjectivs...
%and relatives
%Also info in the Extra module, where more language specific constructions
%can be found.
For example, \verb|FocObj| fronted the object as in sentence \ref{gfSwe:apple}.
\enumsentence{\shortex{6}
{Det & äpplet & vill & jag & inte & ha}
{That & apple & want & I & not &}
{I don't want that apple}}\label{gfSwe:apple}
The module provided functions for preposition stranding:
\enumsentence{\begin{tabular}{ll}
Stranded preposition & cf.\\
Vem måste jag akta mig för? & För vem måste jag akta mig?\\
Who do I need to watch out for? & For whom do I need to watch out?\\
\end{tabular}}

@@

Before describing what has been added, will give a short introductions to the
the cats of resource grammar and the sw implementation in particular.
